---
title: "Introduction to Spatial Cross-Validation"
author: "Mamadou SOW"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Spatial Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Geospatial data presents unique challenges for machine learning that are often overlooked in traditional data science workflows. This vignette explains why spatial cross-validation is essential for reliable model evaluation when working with geographic data.

## The Spatial Dependence Problem

### Tobler's First Law of Geography

> "Everything is related to everything else, but near things are more related than distant things."

This fundamental principle means that spatial observations are rarely independent. Two observations that are close in space tend to have similar characteristics due to:

- **Environmental gradients** (elevation, temperature, precipitation)
- **Spatial processes** (diffusion, spread, influence)
- **Sampling design** (systematic or clustered sampling)
- **Spatial autocorrelation** in the underlying phenomena

### Why Traditional Cross-Validation Fails

Standard random k-fold cross-validation assumes that observations are independent and identically distributed (i.i.d.). When this assumption is violated by spatial dependence:

```{r}
# Example: What happens with random CV on spatial data
set.seed(123)
n <- 100
x <- runif(n, 0, 100)
y <- runif(n, 0, 100)
z <- 10 + 0.5*x + 0.3*y + rnorm(n, 0, 2)  # Spatially structured variable

# Random CV might put nearby points in both train and test
train_idx <- sample(1:n, 80)
test_idx <- setdiff(1:n, train_idx)

# Calculate minimum distance between train and test
distances <- numeric(length(test_idx))
for (i in seq_along(test_idx)) {
  distances[i] <- min(sqrt((x[test_idx[i]] - x[train_idx])^2 + 
                          (y[test_idx[i]] - y[train_idx])^2))
}
min(distances)  # Often very small!
```

### Consequences of Spatial Leakage

When training and test observations are spatially close:

1. **Over-optimistic performance estimates**: The model sees similar patterns during training and testing
2. **Underestimated generalization error**: True spatial generalization is not measured
3. **Misleading model selection**: Models may be chosen based on inflated performance metrics
4. **Poor real-world performance**: Models fail when applied to new geographic areas

## The Solution: Spatial Cross-Validation

Spatial cross-validation methods explicitly control the separation between training and test observations to ensure:

- **Spatial independence** between train and test sets
- **Realistic generalization estimates** across geographic space
- **Reliable model comparison** under spatial constraints
- **Better understanding** of spatial model performance

## When to Use Spatial Cross-Validation

You should use spatial cross-validation when:

- Your data has explicit spatial coordinates
- Observations may be spatially autocorrelated
- You need to predict in new geographic areas
- Spatial patterns are important in your domain
- You want conservative performance estimates

## Common Applications

Spatial cross-validation is particularly important in:

- **Agriculture**: Crop yield prediction, soil mapping
- **Environmental science**: Species distribution modeling, pollution mapping
- **Climatology**: Temperature and precipitation interpolation
- **Hydrology**: Water quality prediction, flood risk assessment
- **Forestry**: Timber volume estimation, disease detection
- **Urban planning**: Property value prediction, crime mapping
- **Remote sensing**: Land cover classification, biomass estimation
- **Epidemiology**: Disease spread modeling, health risk assessment

## What spatialcvR Provides

The `spatialcvR` package offers:

- **Multiple spatial CV methods**: Block, buffered, clustering approaches
- **Spatial leakage detection**: Identify risky train/test proximity
- **Model evaluation metrics**: RMSE, MAE, R², MAPE
- **Spatial residual diagnostics**: Analyze error distribution in space
- **Method comparison**: Compare spatial vs. random CV
- **Flexible input**: Works with sf objects, data frames, or coordinates
- **CRS-aware**: Proper coordinate system handling

## Quick Example

```{r}
library(spatialcvR)

# Load sample data
data(sample_spatial_data)

# Create spatial folds
folds <- spatial_folds(
  data = sample_spatial_data,
  x = "longitude", 
  y = "latitude",
  k = 5,
  method = "block"
)

# Examine the folds
print(folds)

# Detect spatial leakage
leakage <- detect_spatial_leakage(
  data = sample_spatial_data,
  folds = folds,
  x = "longitude",
  y = "latitude"
)

print(leakage)
```

## Next Steps

- Learn about different [spatial cross-validation methods](spatial-cross-validation.html)
- Understand [spatial leakage detection](spatial-leakage.html)
- Explore [model evaluation and comparison](model-evaluation.html)

## References

- Brenning, A. (2012). "Spatial cross-validation and bootstrap for the assessment of prediction rules in remote sensing". 
- Roberts, D. R., et al. (2017). "Cross-validation strategies for data with temporal, spatial, hierarchical, or phylogenetic structure".

## Key Takeaways

1. **Spatial dependence violates i.i.d. assumptions** in traditional cross-validation
2. **Random CV can give misleading results** for geospatial data
3. **Spatial CV methods control train/test separation** for reliable evaluation
4. **Use spatialcvR when working with geographic data** to ensure robust model assessment